---
title: "Trends in Pacific Canadian groundfish stock status"
author:
- "Sean C. Anderson^1^"
- "Other authors here^1^"
date: |
  ^1^Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, V6T 6N7, Canada\
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    includes:
      in_header: "preamble.tex"
fontsize: 12pt
bibliography: "refs.bib"
csl: "csas.csl"
link-citations: no
linkcolor: "black"
urlcolor: "black"
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(here)
opts_chunk$set(echo = FALSE, message = FALSE, warnings = FALSE)

status_dat <- readRDS(here("data-generated/x_t_posterior.rds"))
status <- dplyr::filter(status_dat, year == 2020)
blrp <- dplyr::filter(status, ratio == "B/LRP")
busr <- dplyr::filter(status, ratio == "B/USR")
bbmsy <- dplyr::filter(status, ratio == "B/B[MSY]")

med1 <- function(x) {
  sprintf("%.1f", round(median(x), 1))
}
lwr1 <- function(x) {
  sprintf("%.1f", round(quantile(x, probs = 0.025), 1))
}
upr1 <- function(x) {
  sprintf("%.1f", round(quantile(x, probs = 0.975), 1))
}

p_thresh <- readRDS(here("data-generated/p-thresh.rds"))
```

```{r pcod, include=FALSE}
pcod3cd <- dplyr::filter(p_thresh, stock_clean == "Pacific Cod 3CD") %>%
  pull(p_lrp)
pcod3cd <- sprintf("%.1f", round(pcod3cd * 100, 1))
```

# Highlights

<!-- - In 2020, the West Coast Haida Gwaii synoptic trawl, southern outside hard -->
<!--   bottom long line, International Pacific Halibut Commission, and Sablefish -->
<!--   trap surveys were conducted. The West Coast Vancouver Island synoptic trawl -->
<!--   survey and inside hard bottom long line surveys did not run due to COVID-19. -->
<!--   OR SKIP THIS -->

- In 2020, four of the usual surveys that monitor Pacific Canadian groundfish
  stocks were conducted while two were suspended due to COVID-19.

- We assembled the posterior distributions of biomass status time series for 24
  Pacific Canadian groundfish stocks assessed in the last decade and fit
  a hierarchical model to estimate average, and stock-specific, status until 2020.

- Something about most recent assessments... something about probability
  density below LRP and USR...

- Something about the clear declining trend that was arrested around 2000. Mean
  trend relatively stable since 2000. Coincides with ITQs and synoptic trawl
  surveys.

- Something about considerable variability within and across individual stocks
  around the overall mean; pick examples

# Fisheries-independent groundfish surveys

DFO conducts a suite of randomized surveys using bottom trawl, longline hook, and longline trap gear that, in aggregate, cover Canada's Pacific Coast.
Synoptic trawl surveys in Queen Charlotte Sound and Hecate Strait are conducted in odd numbered years and the West Coast of Vancouver Island (WCVI) and the West Coast of Haida Gwaii (WCHG) surveys are conducted in even numbered years.
In addition to the bottom trawl surveys, four biennial Hard Bottom Longline (HBLL) surveys are conducted, two in "inside" waters (east of Vancouver Island) and two others in "outside" waters (everything else).
Each year the HBLL surveys alternate between northern and southern areas.
Lastly, a coast-wide longline trap survey targeting sablefish is conducted every year and DFO collects biological information from the International Pacific Halibut Commission (IPHC) Setline Survey.
In 2020, the synoptic WCHG trawl survey, outside HBLL south survey, IPHC setline survey, and sablefish survey were conducted.
The synoptic WCVI survey did not run due to COVID-19.
As of writing, the only 2020 data in DFO's survey databases are the WCHG and sablefish surveys.
Because of this, and because of legislative changes described below, we focus our report this year on groundfish biological status from assessments over the last decade.

# A state-space model of groundfish stock status

Recent amendments to the *Fisheries Act* via the Fish Stocks provisions have increased DFO's focus on stock status with respect to biological reference points.
The Precautionary Approach (PA) Framework describes two stock status reference points: (1) the Limit Reference Point (LRP), a "status below which serious harm is occurring to the stock", and (2) the Upper Stock Reference Point (USR), which represents the "threshold below which removals must be progressively reduced in order to avoid reaching the LRP" [@dfo2009PA, Figure \@ref(fig:pa-framework)].
The USR defines the breakpoint between the healthy and cautious zone and the LRP defines the breakpoint between the cautious and critical zone (Figure \@ref(fig:pa-framework)).
The LRP and USR are often calculated based on B~MSY~ (biomass at maximum sustainable yield) at provisional levels of 0.4 and 0.8 B~MSY~, but in other cases are based on historical reference points (historical biomass values that match the definition of the LRP or USR).

Assessment scientists conduct regular stock assessment of major fish stocks in BC.
These assessments combine fishery-dependent data, such as catches, and fishery-independent data (data from scientific surveys), to estimate quantities such as growth, natural mortality, and maturity, and ultimately derive measures of stock status and fishing intensity.
Here, we gather output data from these assessments representing the posterior distribution of biomass compared to three measures of stock status: (1) B/B~MSY~ (biomass divided by B~MSY~), (2) B/USR (biomass divided by the USR), and (3) B/LRP (biomass divided by the LRP).
We develop a Bayesian state-space time-series model to explore trends in these measures of status across stocks until the year 2020.
To do this, we build on a model described in @hilborn2020 in two ways: (1) we incorporate uncertainty in stock-specific status and (2) we project the latent underlying stock-specific status forward to the last year (2020).
The latter accounts for not all stocks having an up-to-date stock assessment and accounts for future possible stock states based on the time-series properties of the fitted stock trajectories.

Using $x_t$ to denote **log** stock status at time $t$, we model $x_t$ as
$$
\begin{aligned}
x_{t=1} &=  x_0 ,\\
x_{t>1} &= x_{t-1} + \eta_t, \;
      \eta_t \sim \mathrm{Normal}
      \left( 0, \sigma_{\eta}^2 \right),
\end{aligned}
$$
where $\eta_t$ is a random walk deviation with standard deviation $\sigma_\eta$.
We assume an auto-regressive observation model:
$$
\begin{aligned}
\hat{y}_{j,t} &\sim \mathrm{Normal} \left( y_{j,t}, \hat{\tau}^2\right),\\
y_{j,t} &= x_{t} + \alpha_j + \epsilon_{j,t}, \;
      \epsilon_{j,t} \sim \mathrm{Normal}
      \left(\rho \epsilon_{j,t-1}, \sigma_{\epsilon}^2 \right),
\end{aligned}
$$
where $\hat{y}_{j,t}$ is the mean log stock status for stock $j$ and time $t$ (from an assessment), $y_{j,t}$ is the unobserved "true" mean stock-specific status, and $\hat{\tau}$ is the standard deviation of the log stock status from an assessment.
We group Pacific Cod and Walleye Pollock stocks to share a $\rho$ and $\sigma_{\epsilon}$ and all other stock to share another $\rho$ and $\sigma_{\epsilon}$ (stock trajectories were considerably more "wiggly" for these species).
The symbol $\alpha_j$ represents a stock-specific intercept that is constrained such that the sum of all $\alpha_j$ is zero to make the overall mean $x_t$ identifiable.
The symbol $\epsilon_{j,t}$ represents a first-order autoregressive (AR1) deviation with correlation $\rho$ and standard deviation $\sigma_\epsilon$.
We place half-Normal(0, 1) priors on $\sigma_{\epsilon}$ and $\sigma_{\eta}$, Normal(0, 5^2^) priors on $x_0$ and $\alpha_j$, and a Normal(0, 1)[-1, 1] prior on $\rho$.
We fit the models with Stan [@carpenter2017; @rstan2020], sampling 1000 iterations on each of 6 chains and discarding the first half as warm up.
Data and code for our analysis are available at \url{https://github.com/pbs-assess/gftrends}.

# Status and trends

We collected assessment biological stock status data from 24 groundfish assessments in BC. Four stocks provided historical reference points for management advice, 18 provided MSY-based reference points for management advice, and two stocks (Rocksole 5AB and 5CD) provided both historical and MSY reference points (here we use the MSY-based ones). The most recent assessments were for Rougheye/Blackspotted Rockfish, Pacific Cod, Bocaccio, inside and outside Yelloweye Rockfish, and Sablefish (Figure \@ref(fig:ridges)).

Estimated biomass was above the LRP and USR for most stocks as of the last assessment (Figure \@ref(fig:ridges)).
Lingcod 4B and Bocaccio 5ABCD were the only stocks (with full posterior distributions available) with $>$ 5% posterior density below their LRP as of their last assessment (Figure \@ref(fig:ridges)).
Pacific Cod in 3CD almost met this threshold with a `r pcod3cd`% probability B $<$ LRP in 2020.
The assessment of Bocaccio, however, predicted the stock to rebuild to XX by XX due to a large recruitment event in XX (REF).
Yelloweye Rockfish in outside BC waters (the one recently assessed stock not included in our dataset) were previously estimated to be below their LRP with high probability [@dfo2015yelloweyeoutside]; however, a recent rebuilding assessment [@cox2020] found the stock is likely above its LRP.
Quillback Rockfish in both outside and inside waters had some posterior density below their LRPs as of the 2011 assessment, but the full posterior distribution was not available and so is not shown in Figure \@ref(fig:ridges) but is shown as quantiles in Figure \@ref(fig:ts-lrp).
<!-- As of their last assessment date, `r sum(p_thresh$p_lrp > 0.05)`/`r nrow(p_thresh)` stocks in Figure \@ref(fig:ridges) had > 5% probability of being below their LRP; -->
Considering the USR instead of LRP, `r sum(p_thresh$p_usr > 0.25)`/`r nrow(p_thresh)` of the stocks in Figure \@ref(fig:ridges) had > 25% probability of being below their USR.

<!-- MSY: -->
<!-- `r sum(p_thresh$p_bbmsy > 0.33, na.rm = TRUE)`/`r sum(!is.na(p_thresh$p_bbmsy))` -->

Across all stocks, there was a clear decline in average biological status for all three indices prior to approximately 2000 (Figure \@ref(fig:overall)).
The late 1990s and early 2000s marked the beginning of a relatively stable average status.
We estimated the overall mean B/LRP in 2020 to be `r med1(blrp$.value)` (95% credible interval [CI]: `r lwr1(blrp$.value)`--`r upr1(blrp$.value)`).
The overall mean B/USR and B/B~MSY~ in 2020 was `r med1(busr$.value)` (95% CI: `r lwr1(busr$.value)`--`r upr1(busr$.value)`) and `r med1(bbmsy$.value)` (95% CI: `r lwr1(bbmsy$.value)`--`r upr1(bbmsy$.value)`) respectively.
Average status uncertainty in the last few years increased due to not all stocks having a current assessment (Figure \@ref(fig:overall)).

The clear pattern in the average biological status has been accompanied by considerable variation within and across individual stocks (Figure \@ref(fig:ts-lrp)).
Within individual stocks, Pacific Cod and Walleye Pollock---both with shorter generation times compared to many other assessed stocks here---exhibited strong patterns of decadal variation.
Across stocks, Redstripe Rockfish in northern and southern BC and coastwide Yellowtail Rockfish are examples of stocks with B/LRP trajectories consistently above the average.
Conversely, Lingcod 4B and Bocaccio 5ABCD are examples of stocks with B/LRP estimated at consistently below the average.
Sablefish followed the average B/LRP trajectory until around 1990 when they continued to decline while the average stabilized.
The Sablefish stock trajectory has shown signs of a turnaround since around 2015 (Figure \@ref(fig:ts-lrp)).

# Factors influencing trends

There are many potential causes for the observed trends with the most obvious and direct being the impact from fishery removals and management interventions.
The transition from declining average B/LRP and B/USR to a relatively stable trajectory coincided with the implementation of individual transferable quotas (ITQs) for the trawl fleet in 1997 [@turris2000] and with the initiation of the current synoptic trawl surveys in 2003 (REF?).
Quota [@dfo2019ifmp], and therefore catches [@anderson2019syn], for many stocks has remained relatively constant over the last two decades.
The decadal patterns in assessed status for Pacific Cod and Walleye Pollock in early decades are largely driven by trends in commercial catches.
ANYTHING ELSE HERE?

There are also potential species interactions and climatic effects on trends in local biomass density of these species.
ANYTHING ABOUT SPECIES INTERACTIONS?
Loss of spawning habitat due to recent oceanographic conditions has been hypothesized to have led to years of low recruitment in some groundfish [e.g., Pacific Cod in nearby Alaskan waters; @laurel2020].
Conversely, other stocks (e.g., Bocaccio REF) have experienced periods of extremely strong recruitment (44 times the average in XX) presumably due to, at least in part, oceanographic conditions.

Underlying these stock-level trends, ongoing work suggests that temperature velocity---the pace a fish would have to move to maintain consistent temperature---may be related to a fine-scale redistribution of groundfish species Canadian Pacific waters [@english2021].
@english2021 found declines in biomass density over the last decade (negative "biomass velocities", or the equivalent of temperature velocity for biomass) in the warmest locations of the coast for approximately one-quarter of the species examined.
However, at the same time, biomass density remained stable or increased in cooler regions for 94% of the species.
This pattern of redistribution has implications for the efficiency of fishing through a potential for temporary climate-induced hyperstability (catch rates that remain high despite population declines) as organisms concentrate in preferred habitats during population decline, and to strengthen or weaken some species interactions.
<!-- Such a redistribution of species biomass has the potential to generate new ecological and fisheries interactions.-->

<!-- On average, and for roughly half the species-maturity combinations, temperature velocity had a more negative effect on biotic velocity in already warm locations than in relatively cool locations. -->
<!-- approximately one quarter of species-maturity combinations experienced declines (negative biotic velocities) in the warmest locations when experiencing maximum warming. -->
<!-- In contrast, locations experiencing minimal warming or cool locations experiencing maximum warming experienced stable or increasing biotic velocities. -->
<!-- Evidence of declines in biomass in the warmest locations and increases in cooler locations suggests a redistribution of species with the potential for new ecological and fisheries interactions. -->

# References

<div id="refs"></div>

\clearpage

\setlength{\parindent}{0in}
\setlength{\leftskip}{0pt}
\setlength{\parskip}{6pt}

# Figures

(ref:fig-pa) Illustration of DFO's Precautionary Approach Framework.
The two vertical lines---the Limit Reference Point (LRP) and Upper Stock Reference (USR)---are the focus of this analysis.
Based on @dfo2009PA.

```{r pa-framework, fig.cap="(ref:fig-pa)", out.width="3.1in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/pa-framework.pdf"))
```

(ref:fig-ridges) Posterior distribution of three measures of stock status.
Stocks are arranged in order of assessment with the most recent assessments at the top; years in the first column represent the year the status represents.
Colours represent the mean B/LRP value.
Stocks with missing data in the B/B~MSY~ column are assessments where historical reference points were chosen instead of MSY-based reference points.
Vertical dashed lines are at values of 1.0 in all columns and also at values of 0.4 and 0.8 in the B/B~MSY~ column (0.4 and 0.8 B~MSY~ are provisional LRP and USR values in the PA Framework).
The x-axis has been square-root transformed to slightly compress high ratio values for visualization.
In a few cases (e.g., Bocaccio) the status represents a one- or two-year projection from the last year of available data.

<!-- TODO: SOME HAVE PROJECTIONS? -->

```{r ridges, fig.cap = "(ref:fig-ridges)", out.width="4.2in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/ridges.pdf"))
```

(ref:fig-overall) Overall mean biomass status ($x_t$) for B/LRP, B/USR, and B/B~MSY~.
Dark lines represent the median trajectory, shaded ribbons represent 95% quantile credible intervals, and thin lines represent 25 illustrative draws from the posterior.

```{r overall, fig.cap = "(ref:fig-overall)", out.width="3.5in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/ts-summary.pdf"))
```

(ref:fig-ts-lrp) Trends in B/LRP for 24 groundfish stocks in BC.
Coloured lines and ribbons represent the individual stocks.
Dark coloured lines and shaded ribbons represent output from stock assessments: trajectories of median B/LRP 95% quantile credible intervals.
Thin lines represent draws from the posterior distribution of $y_{j,t}$ (latent stock-specific mean trends; colours) and $x_t$ (latent overall mean trend; grey).
The overall mean trend ($x_t$; grey) is the same across panels.

```{r ts-lrp, fig.cap = "(ref:fig-ts-lrp)", out.width="5in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/blrp-x-t.pdf"))
```

