---
title: "Trends in Pacific Canadian groundfish stock status"
author:
- "Sean C. Anderson^1^"
- "Other authors here^1^"
date: |
  ^1^Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, V6T 6N7, Canada\
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
fontsize: 12pt
bibliography: "refs.bib"
csl: "csas.csl"
link-citations: no
linkcolor: "black"
urlcolor: "black"
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(here)
opts_chunk$set(echo = FALSE, message = FALSE, warnings = FALSE)

status_dat <- readRDS(here("data-generated/x_t_posterior.rds"))
status <- filter(status_dat, year == 2020)
blrp <- filter(status, ratio == "B/LRP")
busr <- filter(status, ratio == "B/USR")
bbmsy <- filter(status, ratio == "B/B[MSY]")

med1 <- function(x) {
  sprintf("%.1f", round(median(x), 1))
}
lwr1 <- function(x) {
  sprintf("%.1f", round(quantile(x, probs = 0.025), 1))
}
upr1 <- function(x) {
  sprintf("%.1f", round(quantile(x, probs = 0.975), 1))
}

p_thresh <- readRDS(here("data-generated/p-thresh.rds"))
```

# Highlights

- Ongoing surveys, nothing in 2020, double in 2021... this year we look at trends in assessed stock status

- We developed a data set of the full posterior distribution of three measures
  of stock status (...) from 24 stock assessments in BC. These represent... We
  fit a hierarchical time series model to evaluate overall trends and status.

- Something about the posterior distributions in the last assessed year; highlight a few stocks

- Something about the clear declining trend that was arrested around 2000. Mean
  trend relatively stable since 2000, by considerable variability for some
  individual stocks such as ...

- Something about the overall estimated mean status as of 2020

# A state-space time-series model of groundfish stock status

The Groundfish Section at Pacific Biological Station conducts regular stock assessment of...
These assessments take into account...
Here, we develop a Bayesian hierarchical time-series model to explore trends in
overall status across stocks up until the year 2020...

We model three measures of stock status:
(1) B/B~MSY~ (biomass divided by biomass at maximum sustainable yield),
(2) B/USR (biomass divided by the upper stock reference), and
(3) B/LRP (biomass divided by the limit reference point).
To do this, we build on a model described in @hilborn2020 in two ways:
(1) we incorporate uncertainty in stock-specific status and
(2) we project the latent underlying stock-specific status forward to the last year (2020).
The latter accounts for not all stocks having an up-to-date stock assessment
and accounts for future possible stock states based on the time-series properties
of the fitted stock trajectories.

Using $x_t$ to denote **log** stock status at time $t$, we model $x_t$ as
$$
\begin{aligned}
x_{t=1} &=  x_0 ,\\
x_{t>1} &= x_{t-1} + \eta_t, \;
      \eta_t \sim \mathrm{Normal}
      \left( 0, \sigma_{\eta}^2 \right),
\end{aligned}
$$
where $\eta_t$ is a random walk deviation with standard deviation $\sigma_\eta$.
We assume an auto-regressive observation model:
$$
\begin{aligned}
\hat{y}_{j,t} &\sim \mathrm{Normal} \left( y_{j,t}, \hat{\tau}^2\right),\\
y_{j,t} &= x_{t} + \alpha_j + \epsilon_{j,t}, \;
      \epsilon_{j,t} \sim \mathrm{Normal}
      \left(\rho \epsilon_{j,t-1}, \sigma_{\epsilon}^2 \right),
\end{aligned}
$$
where $\hat{y}_{j,t}$ is the mean log stock status for stock $j$
and time $t$ (from an assessment), $y_{j,t}$ is the unobserved "true" mean stock-specific status,
and $\hat{\tau}$ is the standard deviation of the log stock status from an assessment.
We group Pacific Cod and Walleye Pollock stocks to share a $\rho$ and $\sigma_{\epsilon}$
and all other stock to share another $\rho$ and $\sigma_{\epsilon}$ (stock trajectories were considerably more "wiggly" for these species).
The symbol $\alpha_j$ represents a stock-specific intercept
that is constrained such that the sum of all $\alpha_j$ is zero to make the overall mean $x_t$ identifiable.
The symbol $\epsilon_{j,t}$ represents a first-order autoregressive (AR1) deviation
with correlation $\rho$ and standard deviation $\sigma_\epsilon$.
We place half-Normal(0, 1) priors on $\sigma_{\epsilon}$ and $\sigma_{\eta}$,
Normal(0, 5^2^) priors on $x_0$ and $\alpha_j$,
and a Normal(0, 1)[-1, 1] prior on $\rho$.
We fit the models with Stan [@carpenter2017; @rstan2020], sampling 1000 iterations
on each of 6 chains and discarding the first half as warm up.

# Status and trends

Main results here; quick bullets for now

- Lingcod 4B  Bocaccio 5ABCD the only stocks with considerable posterior
  density below the LRP as of the last assessment. Bocaccio, however, are
  predicted to rebuild by XX due to a massive recruitment event in XX (REF).

- Note that Yelloweye Rockfish in outside BC waters were previously estimated
  to be below their LRP [@dfo2015yelloweyeoutside]; an updated assessment (REF)
  found a low probability of this, but is not included here, reasons...
  Quillback, some probability as of XX, but full posterior not available and so
  not shown in Figure \@ref(fig:overall) but is shown in Figure
  \@ref(fig:ts-lrp).

- No clear correlation between year of last assessment and stock status (Figure
  \@ref(fig:overall))

- There was a clear decline in status for all three metrics prior to
  approximately 2000 (Figure \@ref(fig:overall)). Status prior to 2000 is more
  uncertain since there is not a continuous random stratified survey and a lack
  of commercial observers...

- Mean status since about 2000 has been relatively stable, although with
  considerable variation within some stocks (e.g., Pacific Cod, Walleye
  Pollock, and Redstripe Rockfish stocks)

- Overall mean B/LRP is estimated to be `r med1(blrp$.value)` (95% CI: `r lwr1(blrp$.value)`--`r upr1(blrp$.value)`) in 2020, and we note
  the increased uncertainty in the last few years due to not all assessments
  being current (Figure \@ref(fig:overall)).

- Overall mean B/USR and B/B~MSY~ are estimated at `r med1(busr$.value)` (95% CI: `r lwr1(busr$.value)`--`r upr1(busr$.value)`) and `r med1(bbmsy$.value)` (95% CI: `r lwr1(bbmsy$.value)`--`r upr1(bbmsy$.value)`) in 2020

-  `r sum(p_thresh$p_usr > 0.25)`/`r nrow(p_thresh)` in Figure \@ref(fig:ridges) stocks had > 25% probability of being below their USR as of the last
  date of assessment

```{r pcod, include=FALSE}
pcod3cd <- filter(p_thresh, stock_clean == "Pacific Cod 3CD") %>% pull(p_lrp)
pcod3cd <- sprintf("%.1f", round(pcod3cd * 100, 1))
```

- `r sum(p_thresh$p_lrp > 0.05)`/`r nrow(p_thresh)` stocks had > 5% probability of being below their LRP as of the last
  date of assessment. Pacific Cod 3CD: `r pcod3cd`% in 2020

# Factors influencing trends

There are many potential causes for the observed trends with the most obvious
and direct being the impact from fishery removals and management interventions.
There are also potential species interactions and climatic effects on trends in
local biomass density of these species. Ongoing work suggests ... Philina's
work here...

\vspace{6pt}

# References

\vspace{9pt}

<div id="refs"></div>

\clearpage

\setlength{\parindent}{0in}
\setlength{\leftskip}{0pt}
\setlength{\parskip}{6pt}

# Figures

(ref:fig-ridges) Posterior distribution of three measures of stock status.
Stocks are arranged in order of assessment with the most recent assessments at
the top; years in the first column represent the year of the assessment (NOTE
SOME HAVE PROJECTIONS!?). Colours represent the mean B/LRP value. The x-axis
has been square-root transformed two slightly compress high ratio values for
visualization. Stocks with missing data in the B/B~MSY~ column are assessments
where historical reference points were chosen instead of MSY-based reference
points.
Vertical dashed lines are at values of 1.0 in all columns and also at values of 0.4 and 0.8 in the B/B~MSY~ column.

```{r ridges, fig.cap = "(ref:fig-ridges)", out.width="4.2in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/ridges.pdf"))
```

(ref:fig-overall) Overall mean biomass status ($x_t$) for B/LRP, B/USR, and
B/B~MSY~. Dark lines represent the median trajectory, shaded ribbons represent
95% quantile credible intervals, and thin lines represent 25 illustrative draws from the
posterior.

```{r overall, fig.cap = "(ref:fig-overall)", out.width="3.5in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/ts-summary.pdf"))
```


(ref:fig-ts-lrp) Trends in B/LRP for 24 groundfish stocks in BC. Coloured lines
and ribbons represent the individual stocks. Dark coloured lines and shaded ribbons
represent output from stock assessments: trajectories of median B/LRP 95%
quantile credible intervals. Thin lines represent draws from the posterior
distribution of $y_{j,t}$ (latent stock-specific mean trends; colours) and $x_t$
(latent overall mean trend; grey). The overall mean trend ($x_t$; grey) is the same
across panels.

```{r ts-lrp, fig.cap = "(ref:fig-ts-lrp)", out.width="5in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/blrp-x-t.pdf"))
```

(ref:fig-ts-usr) Same as Figure \@ref(fig:ts-lrp) but for B/USR. WILL NEED TO
PICK ONE OF FIGURES 3--5 OR COMBINE THEM SOMEHOW

```{r ts-usr, fig.cap = "(ref:fig-ts-usr)", out.width="5in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/busr-x-t.pdf"))
```

(ref:fig-ts-bbmsy) Same as Figure \@ref(fig:ts-lrp) but for B/B~MSY~. Stocks
with missing data are assessments where historical reference points were chosen
instead of MSY-based reference points.

```{r ts-bbmsy, fig.cap = "(ref:fig-ts-bbmsy)", out.width="5in", fig.align="center", fig.pos="ht"}
include_graphics(here("figs/bbmsy-x-t.pdf"))
```

